Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-05 doc/sc2012/circuit.tex
Lots of little changes

% FIXME   We add here a description of how the colors print when printed greyscale.
%
Figure~\ref{sfig:part_fig:pvs} shows how the nodes
in a small graph might be split into three pieces.  Blue nodes,
attached by wires only to nodes in the same piece, are 
{\em private} to the piece.  Orange nodes, on the boundary of a
piece, are {\em shared} with (connected to) other pieces.
In the simulation, computations on the
private nodes of different pieces are independent, while
computations on the shared nodes require communication.  To make
this explicit in the program, we partition the nodes region into
private and shared subregions (line 18).  To partition a region, we
provide a {\em coloring}, which is a relation between the elements of
a region and a set of colors.  
%A partition is an object which given a
%coloring and a region $r$ contains for each color $c$ a subregion of
%$r$ containing all the elements of $r$ colored $c$. 
For each color $c$ in the coloring, the partition contains a subregion $r$
of the region being partitioned, with $r$ consisting of the elements colored $c$.  Note that the
partition into shared and private nodes is disjoint because each node has one color.

>>>>

% FIXME   We add here a description of how the colors print when printed greyscale.
%
Figure~\ref{sfig:part_fig:pvs} shows how the nodes
in a small graph might be split into three pieces.  Blue (lighter) nodes,
attached by wires only to nodes in the same piece, are 
{\em private} to the piece.  Orange (darker) nodes, on the boundary of a
piece, are {\em shared} with (connected to) other pieces.
In the simulation, computations on the
private nodes of different pieces are independent, while
computations on the shared nodes require communication.  To make
this explicit in the program, we partition the nodes region into
private and shared subregions (line 18).  To partition a region, we
provide a {\em coloring}, which is a relation between the elements of
a region and a set of colors.  
%A partition is an object which given a
%coloring and a region $r$ contains for each color $c$ a subregion of
%$r$ containing all the elements of $r$ colored $c$. 
For each color $c$ in the coloring, the partition contains a subregion $r$
of the region being partitioned, with $r$ consisting of the elements colored $c$.  Note that the
partition into shared and private nodes is disjoint because each node has one color.

### 2012-05-05 doc/sc2012/experiments.tex
Lots of little changes

The circuit simulation has a simple application-specific mapper.  At initialization
time, the mapper queries the list of GPUs in the machine and identifies each GPU's
framebuffer memory and {\em zero-copy} memory (pinned memory that both the GPU and
CPUs in the system can access directly).  Once the circuit is partitioned, the partitions
are assigned a home GPU in round-robin fashion.  Every task related to that partition is
then sent to the home GPU, with no task stealing allowed.  
(In a well-partitioned circuit,
load imbalance is low enough that the cost of moving the private data for a piece from one
GPU to another outweighs any benefits.)

@@

Our second experiment is a port of the \emph{fluidanimate} benchmark
from the PARSEC benchmark suite\cite{bienia11benchmarking}, which does a
particle-based simulation of an incompressible fluid.  Each particle
interacts only with other nearby particles. The benchmark divides the
space in which the fluid can move into a three-dimensional array of
cells such that the range of interaction is limited to just the cells
adjacent (including diagonals) to the one a particle resides in.  The
application divides the array into {\em grids} and assigns each grid to
a thread.  Per-cell locking safely accesses particles in
cells that lie on the edge of a grid.  This fine-grained locking
scheme along with the assumption of a shared address space gives
good scaling in a multi-core processor, but rules out any attempt to run
beyond a single node.

@@

\subsection{Adaptive Mesh Refinement}
\label{subsec:exp_amr}
Our final application is based on the third heat equation example from
the Berkeley Labs BoxLib project \cite{BoxLib}.  This application is a
three-level adaptive-mesh-refinement (AMR) code that computes a first
order stencil on a 2D mesh of cells.  
%The simulation iterates for many time steps.  
Updating the simulation for one time step consists of
three phases.  In the first phase, the boundary cells around a box at
a refined level linearly interpolate their values from the nearby
cells at the next coarsest level.  The second phase performs the
stencil computation on each cell in every level.  In the third phase,
cells at a coarser level that have been refined are restricted to the
average of the cells that they physically contain at the next finest
level of refinement.

Achieving high-performance on this application is particularly
challenging for several reasons.  First, the application has a very
high communication-to-computation ratio which, for a fixed problem
size, begins as being memory bound and with increasing node count
becomes network bound as the perimeter-to-area ratio of cell grids
increases.  Second, when choosing how to partition cells into grids,
the programmer must consider the locality between cells within a
level as well as across levels.  For
cross-level cell dependences, mapping decisions must be made at
runtime as the location of refinements are dynamically determined.
Finally, this application has parallelism both
between tasks running at the same level and tasks running across
levels, leading to complicated input-dependent data dependences.

@@

We compared our Legion implementation against BoxLib on three
different problem sizes with a fixed number of cells per level of
refinement, but with randomly chosen refinement locations.  BoxLib
also supports OpenMP and we took their best performance from using 1,
2, 4, or 8 threads per node.  Our Legion implementation always uses
one thread per node to illustrate that in this application locality is
significantly more important than parallelism.

@@

As the node count increases the AMR code becomes highly dependent on
interconnect performance.  BoxLib performs much better on Keeneland
than on Viz due to the better interconnect.  
At higher node counts BoxLib
begins to catch up (see Figure~\ref{fig:amr_keeneland})
because our intra-level ghost-cell exchange 
uses GASNet memory to exchange ghost cells, requiring a
linear increase in network traffic with the number of nodes.  BoxLib
uses direct node-to-node exchanges of ghost cells, similar to
our fluid application.  A future implementation of our AMR code will
employ a similar ghost cell exchange algorithm to improve scalability.

>>>>

The circuit simulation has a simple application-specific mapper.  At initialization
time, the mapper queries the list of GPUs in the machine and identifies each GPU's
framebuffer memory and {\em zero-copy} memory (pinned memory that both the GPUs and
CPUs on a node can access directly).  Once the circuit is partitioned, the partitions
are assigned a home GPU in round-robin fashion.  Every task related to that partition is
then sent to the home GPU, with no task stealing allowed.  
(In a well-partitioned circuit,
load imbalance is low enough that the cost of moving the private data for a piece from one
GPU to another outweighs any benefits.)

@@

Our second experiment is a port of the \emph{fluidanimate} benchmark
from the PARSEC benchmark suite\cite{bienia11benchmarking}, which does a
particle-based simulation of an incompressible fluid.  Each particle
interacts only with nearby particles. The benchmark divides the
space in which the fluid can move into a three-dimensional array of
cells such that the range of interaction is limited to just the cells
adjacent (including diagonals) to the one a particle resides in.  The
application divides the array into {\em grids} and assigns each grid to
a thread.  Per-particle locking safely accesses particles in
cells that lie on the edge of a grid.  This fine-grained locking
scheme along with the assumption of a shared address space gives
good scaling in a multi-core processor, but prohibits running
beyond a single node.

@@

\subsection{Adaptive Mesh Refinement}
\label{subsec:exp_amr}
Our final application is based on the third heat equation example from
the Berkeley Labs BoxLib project \cite{BoxLib}.  This application is a
three-level adaptive-mesh-refinement (AMR) code that computes a first
order stencil on a 2D mesh of cells.  
%The simulation iterates for many time steps.  
Updating the simulation for one time step consists of
three phases.  In the first phase, the boundary cells around a box at
a refined level linearly interpolate their values from the nearby
cells at the next coarser level.  The second phase performs the
stencil computation on each cell in every level.  In the third phase,
cells at a coarser level that have been refined are restricted to the
average of the cells that they physically contain at the next finest
level of refinement.

Achieving high-performance on this application is particularly
challenging for several reasons.  First, the application has a very
high communication-to-computation ratio which, for a fixed problem
size, begins as being memory bound and with increasing node count
becomes network bound as the perimeter-to-area ratio of cell grids
increases.  Second, when choosing how to partition cells into grids,
the programmer must consider the locality between cells within a
level as well as across levels.  For
cross-level cell dependences, optimal mapping decisions can only be made at
runtime as the location of refinements are dynamically determined.
Finally, this application has parallelism both
between tasks running at the same level and tasks running across
levels, leading to complicated input-dependent data dependences.

@@

We compared our Legion implementation against BoxLib on three
different problem sizes with a fixed number of cells per level of
refinement, but with randomly chosen refinement locations.  BoxLib
also supports OpenMP and we took their best performance from using 1,
2, 4, or 8 threads per node.  Our Legion implementation always uses
one thread per node to illustrate that in this application locality is
significantly more important than fine-grained data-parallelism.

@@

As the node count increases the AMR code becomes highly dependent on
interconnect performance.  BoxLib performs much better on Keeneland
than on Viz due to the better interconnect.  
At higher node counts BoxLib
begins to catch up (see Figure~\ref{fig:amr_keeneland})
because our application's intra-level ghost-cell exchange algorithm
uses GASNet memory to communicate ghost cells, requiring a
linear increase in network traffic with the number of nodes.  BoxLib
uses direct node-to-node exchanges of ghost cells, similar to
our fluid application.  A future implementation of our AMR code will
employ a similar ghost cell exchange algorithm to improve scalability.

### 2012-05-05 doc/sc2012/related.tex
Lots of little changes

Legion began as an outgrowth of Sequoia, a locality-aware programming language capable
of expressing computations in deep memory hierarchies
\cite{Fatahalian06}.  Sequoia is a special case of the Legion
programming model in which only arrays can be recursively partitioned,
all access is exclusive, there is a static mapping of tasks and data (though
extensions to Sequoia make this mapping more dynamic \cite{BCSA11}) and, most fundamentally, the decomposition of tasks and the decomposition
of data is one-to-one.  Legion generalizes the Sequoia model by allowing for
partitioning of pointer data structures through regions, enabling dynamic mappings through
the mapper interface, and allowing different coherence
properties.  Legion's decoupling of the task tree from the region tree leads directly to the scheduling problem solved by our software out-of-order processor for tasks with region arguments.

Chapel has several concepts to support the expression of locality \cite{Chamberlain:Chapel}.
Domains are similar to logical regions in that they describe maps from indexes
to same-typed objects.  Domains can create sub-domains by slicing the index sets from
a parent domain.  Domains are a higher level concept than regions; the domain index sets 
support dimensionality and iterators, whereas  logical regions can only be accessed by pointers.  
Second, the act of creating subdomains in Chapel does not track disjointness information,
making it more challenging for the Chapel compiler or runtime to infer task independence.

>>>>

Legion began as an outgrowth of Sequoia, a locality-aware programming language capable
of expressing computations in deep memory hierarchies
\cite{Fatahalian06}.  Sequoia is a special case of the Legion
programming model in which only arrays can be recursively partitioned,
all access is exclusive, there is a static mapping of tasks and data (though
extensions to Sequoia make this mapping more dynamic \cite{BCSA11}) and, most fundamentally, the decomposition of tasks and the decomposition
of data is one-to-one.  Legion generalizes the Sequoia model by allowing for
dynamic partitioning of pointer data structures through regions, enabling dynamic mappings through
the mapper interface, and allowing different coherence
properties.  Legion's decoupling of the task tree from the region tree leads directly to the scheduling problem solved by our software out-of-order processor for tasks with region arguments.

Chapel has several concepts to support the expression of locality \cite{Chamberlain:Chapel}.
Domains are similar to logical regions in that they describe maps from indexes
to objects.  Domains can create sub-domains by slicing the index sets from
a parent domain.  Domains are a higher level concept than regions; the domain index sets 
support dimensionality and iterators, whereas logical regions can only be accessed by pointers.  
Also, the act of creating subdomains in Chapel does not track disjointness information,
making it more challenging for the Chapel compiler or runtime to infer task independence.
